---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

# Personal activity monitoring data analysis

It is now possible to collect a large amount of data about personal movement using activity monitoring devices. Assignment makes use of these data, which consists of values at 5 minute intervals throughout the day.

```{r, echo=FALSE, message=FALSE, warnings=FALSE}
# importing libraries
library(dplyr)
library(xtable)
library(ggplot2)
```


## Loading and preprocessing the data

```{r}
activity <- read.csv("activity.csv", na.strings = "NA")
activity <- tbl_df(activity)
```

## What is mean total number of steps taken per day?

```{r, results='asis'}
calculate_mean_median <- function(data) {
  # select only rows with not-NA number of steps
  steps_by_date <- tapply(data$steps, data$date, sum, na.rm=TRUE)
  hist(steps_by_date, main="Distribution of steps per day", col="wheat", breaks=15)
  rug(steps_by_date)

  mean_steps_df <- aggregate(data$steps, list(data$date), mean, na.rm=TRUE)
  colnames(mean_steps_df) <- c("date", "mean value")

  median_steps_df <- aggregate(data$steps, list(data$date), median, na.rm=TRUE)
  colnames(median_steps_df) <- c("date", "median value")

  result <- merge(mean_steps_df, median_steps_df)
  result
}
original_stat_results <- calculate_mean_median(activity)
```

#### Table of mean/median steps taken in each day
```{r, results='asis'}
print(xtable(original_stat_results), type="html", include.rownames=FALSE)
```

## What is the average daily activity pattern?
```{r, results="asis"}
by_time_interval <- aggregate(activity$steps, list(activity$interval), mean, na.rm=TRUE)
colnames(by_time_interval) <- c("interval", "steps")
plot(by_time_interval, type="l")
```

#### Interval that contains maximum number of steps on average
```{r}
interval_of_max <- by_time_interval[which.max(by_time_interval$steps),]$interval
```
Interval that contains maximum number of steps (on average) is `r interval_of_max`


## Imputing missing values

#### Number of rows with missing values
```{r}
sum(is.na(activity$steps))
```

#### Imputing average for given interval instead of missing value
Every interval with missing (NA) value is populated with mean value for this interval 
```{r, results="asis"}

# create column of the same length as row number in table 
# but containing replicated mean values
one_day <- 60 * 24 / 5
number_of_days <- dim(activity)[1]/one_day
imputed_values <- rep(by_time_interval[, c("steps")], times=number_of_days)

# append column to table
data <- cbind(activity, imputed_values)

conditional_update <- function(steps, imputed_values) {
  ifelse(is.na(steps), imputed_values, steps)
}
 
with_imputed <- mutate(data, steps = conditional_update(steps, imputed_values))
with_imputed <- select(with_imputed, -imputed_values)
```

#### Comparing to estimates from first part (before imputing data)
Comparison table of mean/median for original data and data with imputed values is displayed
```{r, results='asis'}
updated_stat_results <- calculate_mean_median(with_imputed)
colnames(updated_stat_results) <- c("date", "updated mean", "updated median") 

all_stat_results <- merge(original_stat_results, updated_stat_results)
print(xtable(all_stat_results), type="html", include.rownames=FALSE)
```

## Are there differences in activity patterns between weekdays and weekends?
```{r, results="asis"}
with_imputed <-
  with_imputed %>%
  mutate(date=as.Date(date)) %>%
  mutate(daytype=ifelse(weekdays(date, abbreviate(TRUE)) %in% c("Sat","Sun"),
                        "weekend", "weekday"))

weekend_data <- with_imputed[with_imputed$daytype == 'weekend', ]
weekend_averages <- aggregate(weekend_data$steps, list(weekend_data$interval), mean)
weekend_averages$daytype <- "weekend"
                          
weekday_data <- with_imputed[with_imputed$daytype == 'weekday', ]
weekday_averages <- aggregate(weekday_data$steps, list(weekday_data$interval), mean)
weekday_averages$daytype <- "weekday"

combined_averages <- rbind(weekday_averages, weekend_averages)
colnames(combined_averages) <- c("interval", "steps", "daytype")
```

#### Creating plots
```{r}
g = ggplot(combined_averages, aes(interval, steps, color=factor(daytype)))
g + geom_line() + 
    facet_grid(daytype~.) +
    ggtitle("Comparing activity for weekday/weekend") + 
    theme_bw()
```